Lynott et al. (2014) - Potential error in statistics reporting

### 2017-11-09 manuscript.Rmd
Changed the footnotes back to inline note (still need to fix pagebreak problem in footnotes), fixed +/- symbols

```{r echo=FALSE, fig.width=5, fig.height=7}
#  Plot H0 and H1 regions for different significance tests
source("H0H1_plot.R")
```

Figure caption: Figure 1. Illustration of null hypotheses ($H_0$) and alternative hypotheses ($H_1$) for different types of significance tests. **A)** Equivalence test: Tests the hypothesis ($H_0$) that an effect is smaller than $\Delta_{L}$ *or* larger than $\Delta_{U}$. **B)** NHST: Tests the hypothesis ($H_0$) that an effect is equal to 0. **C)** Minimal effects test: Tests the hypothesis ($H_0$) that an effect is larger than $\Delta_{L}$ *and* smaller than $\Delta_{U}$. **D)** Inferiority test: Tests the hypothesis ($H_0$) that an effect is larger than $\Delta$.

@@

Let us imagine that we consult a teacher about the test, and discover that that grades for this test are set at every 6.25% increase in correct answers (F = 0% to 6.25% ... A+ = 93.75% to 100%). We can decide that we are only interested in test differences so far as they correspond to at least a 1 grade point increase or decrease. Thus, our SESOI becomes a difference in raw scores of 6.25%, or 0.0625.

We then calculate the TOST for a two-sample Welch’s t-test, assuming an alpha of 0.05 (two-sided), and using +/- the SESOI of 0.0625 as our equivalence bounds ($\Delta$).

We find that the TOST is non-significant, t(`r round(Moon$TOST_df, 2)`) = `r round(min(abs(c(Moon$TOST_t1, Moon$TOST_t2))), 2)`, p = `r round(max(Moon$TOST_p1, Moon$TOST_p2), 2)` (figure 2C). Thus, we cannot reject the possibility that the true effect really is larger than our SESOI.^[In this example, the bounds of the TOST was set based on a raw effect of 6.25%. We could have converted this raw effect into a standardized effect (Cohen's d) for the current sample, and we would have gotten the exact same result, only then in Cohen's d ($\Delta = +/- `r round(0.0625 / sqrt((sd.asian^2 + sd.control^2) / 2), 2)`)$. Since this is a replication of a previous study, we could also have chosen to calculate the Cohen's d that corresponds to the raw SESOI in the *original* study, and used this as the bounds for the replication. Now we would no longer be answering a question about a raw effect independent of the samples. Instead, we are asking whether we in the replication can reject standardized effects larger than a certain standardized effect in the original study. Since the standard deviations of the original study were only slightly higher than in the replication in this case, Cohen's d (and thus the bounds) would also be only slightly smaller ($\Delta = +/- `r round(0.0625 / sqrt((0.17^2 + 0.20^2) / 2), 2)`)$. As standard deviations of the original study compared to the replication become larger, the bounds become more narrow.] When results are neither statistically different from zero nor statistically equivalent, there is insufficient data to draw conclusions. Further studies are needed, which can be analyzed using a (small-scale) meta-analysis. The additional data will narrow the confidence interval around the observed effect, allowing us to reject the null, reject the SESOI, or both. Analogous to the large sample sizes needed to detect small effects, a lot of data is needed to reject the SESOI when very narrow bounds are used for the equivalence test (e.g., $\Delta_{L}$ = -0.1 and $\Delta_{U}$ = 0.1).

@@

In this case, it is not really clear what an objective criterion for the smallest effect size of interest could be. However, since this is a replication, we could decide that we do not care about an effect if the difference is smaller than the smallest effect that the original study could detect. This essentially means that we use the critical z value (~1.96 in a two-tailed test with an alpha of 0.05) as our bounds ($\Delta = +/- `r round(qnorm(1-0.05/2),2)`$). To figure out what difference corresponds to a critical z in the original study, we multiply critical z with the standard error:

@@

Kahane et al. find no significant correlation between the perceived wrongness of a utilitarian choice in the classical sacrificial dilemmas and the new "greater good" dilemmas, $r(`r Kahane.N-2`) = `r round(Kahane.r.GG, 2)`$, $p = `r round(Kahane.p, 3)`$ ($N = `r Kahane.N`$^[Study 4 in Kahane et al. (2015) had a final sample size of $N = `r nrow(Kahane.data)`$, but due to missing data in `r sum(is.na(Kahane.data))` case, the correlation reported here is based on a sample of only $N = `r Kahane.N`$.]). They conclude that the classical vignettes fail to capture "true"" utilitarianism, but also grant that this conclusion is dependent on the power of their study and that better-powered future studies might overturn their verdict: "Thus, while we cannot rule out the possibility that such an association could emerge in future studies using an even larger number of subjects or different measures, we submit that, in light of the present results, a robust association between ‘utilitarian’ judgment and genuine concern for the greater good seems extremely unlikely." (p. 206).
This inference --- that the result would be surprising if there was a true effect of a size the study could have detected --- can be formalized with an equivalence test for correlations and bounds set to an effect size the study had reasonable power to detect: With `r Kahane.N` participants, the study had 80% power to detect effects of $r = `r round(Kahane.r.80, 2)`$. Given bounds of $\Delta_{L} = `r -round(Kahane.r.80, 2)`$ and $\Delta_{U} = `r round(Kahane.r.80, 2)`$, we find that the result is indeed statistically equivalent, $r(`r Kahane.N-2`) = `r round(Kahane.r.GG, 2)`$, $p = `r round(max(Kahane$TOST_p1, Kahane$TOST_p2), 3)`$. This means that values smaller than $r = `r -round(Kahane.r.80, 2)`$ and larger than $r = `r round(Kahane.r.80, 2)`$ can be rejected at an alpha level of 5% and used as an initial benchmark which future studies could challenge using larger samples and more narrow equivalence bounds.

@@

As with any statistical test, the usefulness of the result of an equivalence test depends on the question we ask. The question manifests itself in the bounds we set: Is our data surprising if the true effect lies outside of the bounds? If we test against very wide bounds, observing statistical significance can hardly be considered surprising, given that most effects in psychology are small to medium (Hemphill, 2003). Examining the papers citing Lakens (2017), we see some researchers state a SESOI without providing a justification for it (Brown and colleagues, 2017; Schumann and colleagues, 2017)^[In these two cases, one could perhaps guess that benchmark values for a Cohen’s d of “medium” size was chosen. ]. An equivalence test using a SESOI of d = 0.5 might very well answer a question the researchers are interested in (for one possible justification based on minimally important differences, see Norman et al., 2003), but researchers should always provide a detailed justification of their chosen SESOI, given that this is in essence a justification for why they ask a specific question from their data.

>>>>

```{r echo=FALSE, fig.width=5, fig.height=7, fig.cap="Illustration of null hypotheses ($H_0$) and alternative hypotheses ($H_1$) for different types of significance tests. **A)** Equivalence test: Tests the hypothesis ($H_0$) that an effect is smaller than $\\Delta_{L}$ *or* larger than $\\Delta_{U}$. **B)** NHST: Tests the hypothesis ($H_0$) that an effect is equal to 0. **C)** Minimal effects test: Tests the hypothesis ($H_0$) that an effect is larger than $\\Delta_{L}$ *and* smaller than $\\Delta_{U}$. **D)** Inferiority test: Tests the hypothesis ($H_0$) that an effect is larger than $\\Delta$."}

#  Plot H0 and H1 regions for different significance tests
source("H0H1_plot.R")
```

@@

Let us imagine that we consult a teacher about the test, and discover that that grades for this test are set at every 6.25% increase in correct answers (F = 0% to 6.25% ... A+ = 93.75% to 100%). We can decide that we are only interested in test differences so far as they correspond to at least a 1 grade point increase or decrease. Thus, our SESOI becomes a difference in raw scores of 6.25%, or 0.0625.
We then calculate the TOST for a two-sample Welch’s t-test, assuming an alpha of 0.05 (two-sided), and using +/- the SESOI of 0.0625 as our equivalence bounds ($\Delta$).
We find that the TOST is non-significant, t(`r round(Moon$TOST_df, 2)`) = `r round(min(abs(c(Moon$TOST_t1, Moon$TOST_t2))), 2)`, p = `r round(max(Moon$TOST_p1, Moon$TOST_p2), 2)` (figure 2C). Thus, we cannot reject the possibility that the true effect really is larger than our SESOI.^[In this example, the bounds of the TOST was set based on a raw effect of 6.25%. We could have converted this raw effect into a standardized effect (Cohen's d) for the current sample, and we would have gotten the exact same result, only then in Cohen's d ($\Delta = \pm `r round(0.0625 / sqrt((sd.asian^2 + sd.control^2) / 2), 2)`)$. Since this is a replication of a previous study, we could also have chosen to calculate the Cohen's d that corresponds to the raw SESOI in the *original* study, and used this as the bounds for the replication. Now we would no longer be answering a question about a raw effect independent of the samples. Instead, we are asking whether we in the replication can reject standardized effects larger than a certain standardized effect in the original study. Since the standard deviations of the original study were only slightly higher than in the replication in this case, Cohen's d (and thus the bounds) would also be only slightly smaller ($\Delta = \pm `r round(0.0625 / sqrt((0.17^2 + 0.20^2) / 2), 2)`)$. As standard deviations of the original study compared to the replication become larger, the bounds become more narrow.] When results are neither statistically different from zero nor statistically equivalent, there is insufficient data to draw conclusions. Further studies are needed, which can be analyzed using a (small-scale) meta-analysis. The additional data will narrow the confidence interval around the observed effect, allowing us to reject the null, reject the SESOI, or both. Analogous to the large sample sizes needed to detect small effects, a lot of data is needed to reject the SESOI when very narrow bounds are used for the equivalence test (e.g., $\Delta_{L}$ = -0.1 and $\Delta_{U}$ = 0.1).

@@

In this case, it is not really clear what an objective criterion for the smallest effect size of interest could be. However, since this is a replication, we could decide that we do not care about an effect if the difference is smaller than the smallest effect that the original study could detect. This essentially means that we use the critical z value (~1.96 in a two-tailed test with an alpha of 0.05) as our bounds ($\Delta = \pm `r round(qnorm(1-0.05/2),2)`$). To figure out what difference corresponds to a critical z in the original study, we multiply critical z with the standard error:

@@

Kahane et al. find no significant correlation between the perceived wrongness of a utilitarian choice in the classical sacrificial dilemmas and the new "greater good" dilemmas, $r(`r Kahane.N-2`) = `r round(Kahane.r.GG, 2)`$, $p = `r round(Kahane.p, 3)`$ ($N = `r Kahane.N`$^[Study 4 in Kahane et al. (2015) had a final sample size of $N = `r nrow(Kahane.data)`$, but due to missing data in `r sum(is.na(Kahane.data))` case, the correlation reported here is based on a sample of $N = `r Kahane.N`$.]). They conclude that the classical vignettes fail to capture "true"" utilitarianism, but also grant that this conclusion is dependent on the power of their study and that better-powered future studies might overturn their verdict: "Thus, while we cannot rule out the possibility that such an association could emerge in future studies using an even larger number of subjects or different measures, we submit that, in light of the present results, a robust association between ‘utilitarian’ judgment and genuine concern for the greater good seems extremely unlikely." (p. 206).
This inference --- that the result would be surprising if there was a true effect of a size the study could have detected --- can be formalized with an equivalence test for correlations and bounds set to an effect size the study had reasonable power to detect: With `r Kahane.N` participants, the study had 80% power to detect effects of $r = `r round(Kahane.r.80, 2)`$. Given bounds of $\Delta_{L} = `r -round(Kahane.r.80, 2)`$ and $\Delta_{U} = `r round(Kahane.r.80, 2)`$, we find that the result is indeed statistically equivalent, $r(`r Kahane.N-2`) = `r round(Kahane.r.GG, 2)`$, $p = `r round(max(Kahane$TOST_p1, Kahane$TOST_p2), 3)`$. This means that values smaller than $r = `r -round(Kahane.r.80, 2)`$ and larger than $r = `r round(Kahane.r.80, 2)`$ can be rejected at an alpha level of 5% and used as an initial benchmark which future studies could challenge using larger samples and more narrow equivalence bounds.

@@

As with any statistical test, the usefulness of the result of an equivalence test depends on the question we ask. The question manifests itself in the bounds we set: Is our data surprising if the true effect lies outside of the bounds? If we test against very wide bounds, observing statistical significance can hardly be considered surprising, given that most effects in psychology are small to medium (Hemphill, 2003). Examining the papers citing Lakens (2017), we see some researchers state a SESOI without providing a justification for it (Brown and colleagues, 2017; Schumann and colleagues, 2017)^[In these two cases, one could perhaps guess that benchmark values for a Cohen’s d of “medium” size was chosen.]. An equivalence test using a SESOI of d = 0.5 might very well answer a question the researchers are interested in (for one possible justification based on minimally important differences, see Norman et al., 2003), but researchers should always provide a detailed justification of their chosen SESOI, given that this is in essence a justification for why they ask a specific question from their data.
